Communicating causal effect heterogeneity

### 2024-07-04 index.qmd
draft figures describing descriptor construction

---
title: Communicating causal effect heterogeneity
running-head: Communicating heterogeneity
author: 
  - name: Matti Vuorre
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    corresponding: true
    affiliation: 
      - ref: 1
  - name: Matthew Kay
    orcid: 0000-0001-9446-0419
    affiliation:
      - ref: 2
  - name: Niall Bolger
    orcid: 0000-0001-8698-8117
    affiliation:
      - ref: 3
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: Northwestern University
    department: Computer Science and Communication Studies
  - id: 3
    name: Columbia University
    department: Department of Psychology
abstract: |
  Advances in experimental, data collection, and analysis methods have brought population variability in psychological phenomena to the fore. Yet, current practices for interpreting such heterogeneity do not appropriately treat the uncertainty inevitable in any statistical summary. Heterogeneity is best thought of as a distribution of features with a mean (average person’s effect) and variance (between-person differences). This expected heterogeneity distribution can be further summarized e.g. as a heterogeneity interval [@bolgerCausalProcessesPsychology2019]. However, because empirical studies estimate the underlying mean and variance parameters with uncertainty, the expected distribution and interval will underestimate the actual range of plausible effects in the population. Using Bayesian hierarchical models, and with the aid of empirical datasets from social and cognitive psychology, we provide a walk-through of effective heterogeneity reporting and display tools that appropriately convey measures of uncertainty. We cover interval, proportion, and ratio measures of heterogeneity and their estimation and interpretation. These tools can be a spur to theory building, allowing researchers to widen their focus from population averages to population heterogeneity in psychological phenomena.
keywords: [heterogeneity, uncertainty, variation, hierarchical model, statistics]
authornote: |
  This manuscript is not yet peer-reviewed.
bibliography: bibliography.bib
citeproc: true
csl: apa.csl
---

@@

m1_parnames <- c(
  "$\\beta_0$", "$\\beta_1$",
  "$\\tau_0$", "$\\tau_1$",
  "$\\rho$", "$\\sigma$"
)
mu1n <- tab_1[2, 2, TRUE]
mu1n_low <- tab_1[2, "CI_low", TRUE]
sd1n <- tab_1[4, 2, TRUE]
sd1n_low <- tab_1[4, "CI_low", TRUE]
mu1 <- number(
  as.numeric(tab_1[2, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)
sd1 <- number(
  as.numeric(tab_1[4, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)
```

@@

```{r}
#| label: fig-1
#| fig-height: 2
#| fig-width: 7
#| fig-cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. A. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. B. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. C. Proportion of negative valence effects in the population (dark). D. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

fig_0 <- ggplot() +
  aes(xdist = dist_normal(mu1n, sd1n)) +
  scale_x_continuous(breaks = extended_breaks()) +
  coord_cartesian(xlim = c(-0.55, 0.25)) +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(lineheight = rel(0.5))
  ) +
  scale_fill_manual(values = c("grey90", "grey80"), na.value = "white")
d1 <- dist_normal(mu1n, sd1n) |>
  mean_qi(.width = c(.66, .9))

fig_1 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(level)),
    .width = c(.68, .95),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$\\tau_1$"),
    x = mu1n + .05, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n + sd1n, y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  geom_text(
    label = TeX("$\\beta_1$"),
    x = mu1n - .075, y = .075,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n, y = .1, yend = 0,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  theme(axis.title.y = element_text(angle = 90))

fig_2 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(level)),
    .width = c(.90, 1),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$HI_{90}$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = d1$.lower[2], xend = d1$.upper[2], y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed", ends = "both")
  )

fig_3 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(x < 0)),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$p^-$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  )

fig_4 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(between(x, -.1, .1))),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$p^{ROPE}$"),
    x = 0.0, y = 0.5, hjust = 0,
    size = 3,
    parse = TRUE
  )

(fig_1 | fig_2 | fig_3 | fig_4) &
  scale_y_continuous(
    expand = expansion(c(0.0, 0.05))
  ) &
  labs(
    y = "Density",
    x = "Valence effect"
  ) &
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  ) &
  plot_annotation(tag_levels = "A")
```

@@

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use .05 and .95, which together define the central 90% of the distribution. Then, we pass those limits and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to get the interval: $HI_{90} = \Phi^{-1}([.05, .95], \beta_1, \tau_1) = \Phi^{-1}([.05, .95],$ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this function calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall in the [`r hi90`] interval. We illustrate this interval in @fig-1 B.

@@

To answer, we pass zero (the critical value) and the estimated mean and standard deviation to the normal cumulative distribution function ($\Phi$; `pnorm()` in R): $p^- = \Phi(0, \beta_1, \tau_1) = \Phi(0,$ `r mu1[1]`, `r sd1[1]`) = `r p_less_zero`. This number is the probability that a random slope from this population would take a negative value, or, in other words, the proportion of individuals in the population with negative valence effects. We illustrate this probability in @fig-1 C.

@@

For example, let us imagine that a theory states that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = \Phi(0.1, \beta_1, \tau_1) - \Phi(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval's theoretical validity. We visualize this probability in @fig-1 D.

@@

post1 <- as_draws_df(
  fit_1, 
  variable = c("b_", "sd_", "cor_", "sigma"), 
  regex = TRUE
)
```

@@

set.seed(99)
x <- seq(-0.6, 0.3, length = 301)

@@

post1_density_mean <- post1 |>
  mean_qi(b_valence1, sd_person__valence1) |>
  calculate_density(b_valence1, sd_person__valence1)

post1_density_samples <- post1 |>
  slice_sample(n = 100) |>
  calculate_density(b_valence1, sd_person__valence1)

@@

```{r}
#| label: fig-2
#| fig-width: 5.4
#| fig-height: 2.2
#| fig-cap: "Bayesian estimates of the heterogeneity distribution of valence effects. A: Probability density function (PDF) curves. The thick line is the same expected PDF of valence effects from Figure 1. Thin lines show 100 PDFs calculated from random draws of $\\beta_1$ and $\\tau_1$ that collectively illustrate the uncertainty in the distribution's location and spread. B: Cumulative density function (CDF) curves, annotated as A. C. Histogram of (draws from) the posterior distribution of $p^-$, which correspond to CDF segments below zero: This value describes the proportion of individuals with negative valence effects on logRT. Point and interval represents the posterior mean and 95\\%CI. Solid and dotted vertical lines in A and B highlight x-axis values of 0, and -0.1 and 0.1, respectively."

# Shared Y axis
scale_y_01 <- scale_y_continuous(
  breaks = extended_breaks(7),
  expand = expansion(c(0.005, 0.005)),
  limits = c(0, 1)
)

# Posterior samples of pdf and cdf curves
p_cdf_samples <- post1_density_mean |>
  ggplot() +
  aes(x, cdf) +
  ylab("Cumulative probability") +
  scale_x_continuous(
    "Valence effect",
    expand = expansion(0),
    breaks = extended_breaks()
  ) +
  scale_y_01 +
  geom_vline(
    xintercept = 0, 
    linewidth = .25
  ) +
  geom_vline(
    xintercept = c(-0.1, 0.1), 
    lty = "dotted", 
    linewidth = .2
  ) +
  geom_line(
    data = post1_density_samples,
    aes(group = .draw),
    alpha = .33,
    linewidth = .15,
    col = "dodgerblue1"
  ) +
  geom_line(
    linewidth = .8,
    col = "dodgerblue4"
  ) +
  theme(legend.position = "none")

p_pdf_samples <- p_cdf_samples +
  aes(y = pdf) +
  scale_y_continuous(
    breaks = extended_breaks(7),
    expand = expansion(c(0.005, 0.005)),
  ) +
  ylab("Probability density")

# Add summaries of CDF < 0 to CDF curves as inset plots
p_p0_samples <- post1 |>
  mutate(dist = dist_normal(b_valence1, sd_person__valence1)) |>
  ggplot() +
  scale_x_continuous(
    expand = expansion(c(0.0, 0.1))
  ) +
  scale_y_01 +
  theme_void(base_size = 9) +
  stat_histinterval(
    aes(y = cdf(dist, 0)),
    breaks = 50,
    fill = alpha("dodgerblue1", 0.5),
    color = "dodgerblue4",
    side = "right",
    .width = .95,
    size = 1,
    scale = 0.5,
    justification = 0.025
  )

p_cdf_samples <- p_cdf_samples +
  inset_element(
    p_p0_samples,
    0, 0, .3, 1, 
    align_to = "panel", 
    on_top = TRUE
  ) +
  theme(plot.tag.position = c(0.25, 0.75))

(p_pdf_samples | p_cdf_samples) +
  plot_annotation(tag_levels = "A")
```

@@

Therefore, in @fig-2 B we depict the heterogeneity distribution as *cumulative distribution* function (CDF) curves based on 100 random posterior draws, together with the mean CDF in a darker color. We believe the CDF is a useful visualization tool because the y-axis describes a directly interpretable quantity: The proportion of the population with valence effects below some specific value.

@@

@@

>>>>

---
title: Communicating causal effect heterogeneity
running-head: Communicating heterogeneity
author: 
  - name: Matti Vuorre
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    corresponding: true
    affiliation: 
      - ref: 1
  - name: Matthew Kay
    orcid: 0000-0001-9446-0419
    affiliation:
      - ref: 2
  - name: Niall Bolger
    orcid: 0000-0001-8698-8117
    affiliation:
      - ref: 3
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: Northwestern University
    department: Computer Science and Communication Studies
  - id: 3
    name: Columbia University
    department: Department of Psychology
abstract: |
  Advances in experimental, data collection, and analysis methods have brought population variability in psychological phenomena to the fore. Yet, current practices for interpreting such heterogeneity do not appropriately treat the uncertainty inevitable in any statistical summary. Heterogeneity is best thought of as a distribution of features with a mean (average person’s effect) and variance (between-person differences). This expected heterogeneity distribution can be further summarized e.g. as a heterogeneity interval [@bolgerCausalProcessesPsychology2019]. However, because empirical studies estimate the underlying mean and variance parameters with uncertainty, the expected distribution and interval will underestimate the actual range of plausible effects in the population. Using Bayesian hierarchical models, and with the aid of empirical datasets from social and cognitive psychology, we provide a walk-through of effective heterogeneity reporting and display tools that appropriately convey measures of uncertainty. We cover interval, proportion, and ratio measures of heterogeneity and their estimation and interpretation. These tools can be a spur to theory building, allowing researchers to widen their focus from population averages to population heterogeneity in psychological phenomena.
keywords: [heterogeneity, uncertainty, variation, hierarchical model, statistics]
authornote: |
  This manuscript is not yet peer-reviewed.
bibliography: bibliography.bib
citeproc: true
csl: apa.csl
editor_options: 
  chunk_output_type: inline
---

@@

m1_parnames <- c(
  "$\\beta_0$", "$\\beta_1$",
  "$\\tau_0$", "$\\tau_1$",
  "$\\rho$", "$\\sigma$"
)
mu1n <- tab_1[2, 2, TRUE]
mu1n_low <- tab_1[2, "CI_low", TRUE]
sd1n <- tab_1[4, 2, TRUE]
sd1n_low <- tab_1[4, "CI_low", TRUE]
mu1 <- number(
  as.numeric(tab_1[2, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)
sd1 <- number(
  as.numeric(tab_1[4, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)

dist_ml <- dist_normal(mu1n, sd1n)
```

@@

```{r}
#| label: fig-1
#| fig-height: 2
#| fig-width: 7
#| fig-cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. A. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. B. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the blue shaded area. C. Proportion of negative valence effects in the population (blue). D. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; blue).

# annotation definitions
valence_dist_color = "gray25"
valence_dist_samples_color = "gray65"
desc_color = "dodgerblue1"
desc_color_dark = "dodgerblue4"
zero_color = alpha("black", 0.75)

arrowhead <- \(...) arrow(length = unit(4, "pt"), type = "closed", ...)

fig_1 <- ggplot() +
  aes(xdist = dist_normal(mu1n, sd1n)) +
  scale_x_continuous(breaks = extended_breaks()) +
  coord_cartesian(xlim = c(-0.55, 0.25)) +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(lineheight = rel(0.5))
  )

d1 <- dist_normal(mu1n, sd1n) |>
  mean_qi(.width = c(.66, .9))

fig_1_params <- fig_1 +
  stat_slab(
    aes(fill = after_stat(level)),
    .width = c(.68, .95),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$\\tau_1$"),
    x = mu1n + .05, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n + sd1n, y = .1, yend = .1,
    linewidth = .35,
    arrow = arrowhead()
  ) +
  geom_text(
    label = TeX("$\\beta_1$"),
    x = mu1n - .075, y = .075,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n, y = .1, yend = 0,
    linewidth = .35,
    arrow = arrowhead()
  ) +
  scale_fill_manual(values = c("grey90", "grey80"), na.value = "white") +
  theme(axis.title.y = element_text(angle = 90))

fig_1_descriptors <- fig_1 +
  scale_fill_manual(values = c("grey90", alpha(desc_color, 0.45)), na.value = "white")

fig_1_hi <- fig_1_descriptors +
  stat_slab(
    aes(fill = after_stat(level)),
    .width = c(.90, 1),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$HI_{90}$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = d1$.lower[2], xend = d1$.upper[2], y = .1, yend = .1,
    linewidth = .35,
    arrow = arrowhead(ends = "both")
  )

fig_1_p_neg <- fig_1_descriptors +
  stat_slab(
    aes(fill = after_stat(x < 0)),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$p^-$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  )

fig_1_p_rope <- fig_1_descriptors +
  stat_slab(
    aes(fill = after_stat(between(x, -.1, .1))),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$p^{ROPE}$"),
    x = 0.0, y = 0.5, hjust = 0,
    size = 3.5,
    parse = TRUE
  )

(fig_1_params | fig_1_hi | fig_1_p_neg | fig_1_p_rope) &
  scale_y_continuous(
    expand = expansion(c(0.0, 0.05))
  ) &
  labs(
    y = "Density",
    x = "Valence effect"
  ) &
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  ) &
  plot_annotation(tag_levels = "A")
```

@@

```{r}
#| label: fig-hi-point
#| fig-width: 2.7
#| fig-height: 2.2
#| fig-cap: "Construction of point estimates of the limits of the heterogeneity interval. To construct a $\\pi%$ heterogeneity interval, $HI_\\pi$, we use the inverse cumulative distribution function (CDF) of the heterogeneity distribution, $\\Phi^{-1}$. When $\\pi = 0.90$, we have $HI_90 = \\Phi^{-1}\\left((1 \\pm \\pi)/2 | \\beta_1, \\tau_1 \\right) = \\Phi^{-1}\\left([0.05, 0.95] | \\beta_1, \\tau_1 \\right)$."

# Shared Y axis
scale_y_01 <- scale_y_continuous(
  breaks = breaks_extended(7),
  expand = expansion(c(0.005, 0.005)),
  limits = c(0, 1)
)

# shared x axis
xmin <- -0.6
xmax <- 0.3
scale_x_valence <- \(..., breaks = extended_breaks()) scale_x_continuous(
  ...,
  limits = c(xmin, xmax),
  expand = expansion(0),
  breaks = breaks
)

# ML estimate of CDF
p_ml_cdf <-
  ggplot() +
  aes(xdist = dist_ml) +
  stat_slab(
    aes(thickness = after_stat(thickness(cdf))),
    scale = 1,
    linewidth = .8,
    color = valence_dist_color,
    fill = NA
  ) +
  geom_vline(
    xintercept = 0, 
    linewidth = .25,
    color = zero_color,
    linetype = "dashed"
  ) +
  theme(legend.position = "none") +
  scale_y_01 +
  xlab("Valence effect") +
  ylab(quote(Pr("Valence effect" <= italic(x))))

# construction of HI_90 on the CDF
p_ml_cdf_hi <- p_ml_cdf +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_ml, 0.05), y = 0.05, yend = 0.05),
    color = desc_color
  ) +
  geom_segment(
    aes(x = quantile(dist_ml, 0.05), xend = quantile(dist_ml, 0.05), y = 0.05, yend = 0),
    color = desc_color
  ) +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_ml, 0.95), y = 0.95, yend = 0.95),
    color = desc_color
  ) +
  geom_segment(
    aes(x = quantile(dist_ml, 0.95), xend = quantile(dist_ml, 0.95), y = 0.95, yend = 0),
    color = desc_color
  ) +
  geom_segment(
    aes(x = quantile(dist_ml, 0.05), xend = quantile(dist_ml, 0.95), y = 0.025, yend = 0.025),
    color = desc_color,
    arrow = arrowhead(ends = "both")
  ) +
  geom_text(
    aes(x = mu1n, y = .04),
    label = expression(bold(HI)[bold("90")]),
    data = data.frame()[1,],
    size = 3,
    parse = TRUE,
    color = desc_color,
    vjust = 0
  ) +
  annotate("text",
    x = xmin,
    y = c(0.05, 0.95),
    vjust = c(0.3, 0.7),
    label = c(0.05, 0.95),
    color = desc_color,
    hjust = 1.1,
    fontface = "bold",
    size = 2.5
  ) +
  scale_x_valence() +
  coord_cartesian(clip = "off", ylim = c(0,1))

p_ml_cdf_hi
```

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use .05 and .95, which together define the central 90% of the distribution. Then, we pass those limits and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to get the interval: $HI_{90} = \Phi^{-1}([.05, .95] | \beta_1, \tau_1) = \Phi^{-1}([.05, .95] | $ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this function calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall in the [`r hi90`] interval. We illustrate this interval in @fig-1 B and its construction in @fig-hi-point.

@@

```{r}
#| label: fig-p-neg-point
#| fig-width: 2.7
#| fig-height: 2.2
#| fig-cap: "Construction of the point estimate of $p^{-1}$, the proportion of individuals in the population with negative effects. To construct $p^{-1}$, we pass 0 to the cumulative distribution function (CDF) of the heterogeneity distribution."

# construction of p^- on the CDF
p_ml_cdf_p_neg <- p_ml_cdf +
  geom_segment(
    aes(x = 0, xend = 0, y = 0, yend = cdf(dist_ml, 0)),
    color = desc_color
  ) +
  geom_segment(
    aes(x = 0, xend = I(0), y = cdf(dist_ml, 0)),
    color = desc_color,
    arrow = arrowhead()
  ) + 
  annotate("text", 
    x = 0, y = I(0),
    label = "0", 
    color = desc_color, 
    vjust = 1.3, 
    fontface = "bold", 
    size = 3.5
  ) +
  geom_text( 
    aes(x = I(0), y = cdf(dist_ml, 0)),
    label = expression(bolditalic(p ^ "−")), 
    color = desc_color, 
    hjust = 1.2, 
    fontface = "bold", 
    size = 3.5
  ) +
  scale_x_valence(breaks = c(-0.4, -0.2, 0.2)) +
  coord_cartesian(clip = "off", ylim = c(0,1))

p_ml_cdf_p_neg
```

To answer, we pass zero (the critical value) and the estimated mean and standard deviation to the normal cumulative distribution function ($\Phi$; `pnorm()` in R): $p^- = Pr(\textrm{Valence effect} \le 0) = \Phi(0 | \beta_1, \tau_1) = \Phi(0 | $ `r mu1[1]`, `r sd1[1]`) = `r p_less_zero`. This number is the probability that a random slope from this population would take a negative value, or, in other words, the proportion of individuals in the population with negative valence effects. We illustrate this probability in @fig-1 C and its construction using the CDF in @fig-p-neg-point.

@@

For example, let us imagine that a theory states that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = Pr(-0.1 \le \textrm{Valence effect} \le 0.1) = \Phi(0.1 | \beta_1, \tau_1) - \Phi(-0.1 | \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval's theoretical validity. We visualize this probability in @fig-1 D.

@@

post1 <- as_draws_df(
  fit_1, 
  variable = c("b_", "sd_", "cor_", "sigma"), 
  regex = TRUE
) |>
  mutate(dist_valence = dist_normal(b_valence1, sd_person__valence1))
```

@@

set.seed(99)
x <- seq(xmin, xmax, length = 301)

@@

post1_mean <- post1 |>
  mean_qi(b_valence1, sd_person__valence1) |>
  mutate(dist_valence = dist_normal(b_valence1, sd_person__valence1))

post1_density_mean <- post1_mean |>
  calculate_density(b_valence1, sd_person__valence1)

post1_samples <- post1 |>
  slice_sample(n = 100)

post1_density_samples <- post1_samples |>
  calculate_density(b_valence1, sd_person__valence1)

@@

```{r}
#| label: fig-2
#| fig-width: 5.4
#| fig-height: 2.2
#| fig-cap: "Bayesian estimates of the heterogeneity distribution of valence effects. A: Probability density function (PDF) curves. The thick line is the same expected PDF of valence effects from Figure 1. Thin lines show 100 PDFs calculated from random draws of $\\beta_1$ and $\\tau_1$ that collectively illustrate the uncertainty in the distribution's location and spread. B: Cumulative distribution function (CDF) curves, annotated as in A. The blue lines show the calculation of the point estimate of $p^- = Pr(\\textrm{Valence effect} \\le 0)$, the proportion of individuals with negative valence effects on logRT, using the mean CDF of valence effects."

# Posterior samples of pdf and cdf curves
p_samples <- post1_density_mean |>
  ggplot() +
  aes(x) +
  geom_line(
    data = post1_density_samples,
    aes(group = .draw),
    alpha = .3,
    linewidth = .15,
    color = valence_dist_samples_color
  ) +
  geom_line(
    linewidth = .8,
    col = valence_dist_color
  ) +
  geom_vline(
    xintercept = 0, 
    linewidth = .25,
    color = zero_color,
    linetype = "dashed"
  ) +
  theme(legend.position = "none") +
  xlab("Valence effect")

p_cdf_samples <- p_samples +
  aes(y = cdf) +
  scale_x_valence() +
  scale_y_01 +
  ylab(quote(Pr("Valence effect" <= italic(x))))

p_pdf_samples <- p_samples +
  aes(y = pdf) +
  scale_x_valence() +
  scale_y_continuous(
    breaks = extended_breaks(7),
    expand = expansion(c(0.005, 0.005)),
  ) +
  ylab("Probability density")

p_pdf_samples + p_cdf_samples +
  plot_annotation(tag_levels = "A")
```

@@

Therefore, in @fig-2 B we depict the heterogeneity distribution as *cumulative distribution function* (CDF) curves based on 100 random posterior draws, together with the mean CDF in a darker color. We believe the CDF is a useful visualization tool because the y-axis describes a directly interpretable quantity: The proportion of the population with valence effects below some specific value.

@@

```{r}
#| label: fig-hi-samples
#| fig-width: 2.7
#| fig-height: 2.2
#| fig-cap: "Bayesian derivation of the posterior distribution for the lower and upper limits of the heterogeneity interval. Thin gray lines show 100 CDFs of the distribution of valence effects calculated from random draws of $\\beta_1$ and $\\tau_1$, as in @fig-2 B. Blue lines show the calculation of the posterior distribution of $HI_{90} = \\left[\\Phi^{-1}(0.05 | \\beta_1, \\tau_1), \\Phi^{-1}(0.95 | \\beta_1, \\tau_1)\\right]$, which is also depicted as marginal histograms. Compare to the point estimate of the same interval endpoints in @fig-hi-point."

# Add summaries of CDF < 0 to CDF curves as inset plots
p_cdf_intervals <- p_cdf_samples +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_valence, 0.05), y = 0.05, yend = 0.05),
    data = post1_samples,
    color = desc_color,
    alpha = 0.05
  ) +
  geom_segment(
    aes(x = quantile(dist_valence, 0.05), xend = quantile(dist_valence, 0.05), y = 0.05, yend = 0),
    data = post1_samples,
    color = desc_color,
    alpha = 0.1
  ) +
  geom_segment(
    aes(x = I(0), xend = quantile(dist_valence, 0.95), y = 0.95, yend = 0.95),
    data = post1_samples,
    color = desc_color,
    alpha = 0.05
  ) +
  geom_segment(
    aes(x = quantile(dist_valence, 0.95), xend = quantile(dist_valence, 0.95), y = 0.95, yend = 0),
    data = post1_samples,
    color = desc_color,
    alpha = 0.1
  ) +
  stat_histinterval(
    aes(x = quantile(dist_valence, 0.05)),
    y = 0,
    data = post1,
    breaks = 50,
    fill = alpha(desc_color, 0.35),
    color = desc_color_dark,
    side = "bottom",
    .width = .95,
    scale = 1,
    height = 0.1
  ) +
  stat_histinterval(
    aes(x = quantile(dist_valence, 0.95)),
    y = 0,
    data = post1,
    breaks = 50,
    fill = alpha(desc_color, 0.35),
    color = desc_color_dark,
    side = "bottom",
    .width = .95,
    scale = 1,
    height = 0.1
  ) +
  annotate("text",
    x = min(x),
    y = c(0.05, 0.95),
    vjust = c(0.3, 0.7),
    label = c(0.05, 0.95),
    color = desc_color,
    hjust = 1.1,
    fontface = "bold",
    size = 2.5
  ) +
  scale_x_valence() +
  scale_y_continuous(
    breaks = breaks_extended(7),
    expand = expansion(0)
  ) +
  coord_cartesian(clip = "off", ylim = c(0,1))

p_cdf_intervals
```

```{r}
#| label: fig-rope-densities
#| fig-width: 5.4
#| fig-height: 2.8
#| fig-cap: "Bayesian derivation of the posterior distribution for $p^{ROPE}$."

post1_samples_rope <- post1_samples |> 
  slice_head(n = 25)

p_rope_densities <- post1_samples_rope |> 
  ggplot() + 
  stat_slab(
    aes(y = factor(.draw), xdist = dist_valence, fill = after_stat(-0.1 < x & x < 0.1)),
    color = valence_dist_samples_color,
    alpha = 0.5,
    scale = 2.5,
    linewidth = 0.5
  ) +
  geom_vline(xintercept = c(-0.1, 0.1), zero_color, alpha = 0.1, linetype = "22") +
  scale_fill_manual(values = c("TRUE" = desc_color, "FALSE" = alpha(valence_dist_samples_color, 0.3))) +
  guides(fill = "none") +
  coord_cartesian(clip = "off") +
  scale_x_valence() +
  scale_y_discrete(labels = NULL) +
  theme(axis.ticks.y = element_blank()) +
  ylab("Posterior draw") +
  xlab("Valence effect")

p_rope_props <- post1_samples_rope |> 
  slice_head(n = 30) |> 
  ggplot() + 
  aes(y = factor(.draw), x = cdf(dist_valence, 0.1) - cdf(dist_valence, -0.1)) + 
  geom_segment(aes(xend = 0, yend = factor(.draw)), color = desc_color) + 
  geom_point(color = desc_color, size = 0.75) + 
  scale_x_continuous(limits = c(0,1), expand = expansion(0)) +
  scale_y_discrete(labels = NULL) +
  theme(axis.ticks.y = element_blank()) +
  ylab(NULL) +
  xlab(quote(italic(p)^italic(ROPE))) +
  stat_histinterval(
    aes(x = cdf(dist_valence, 0.1) - cdf(dist_valence, -0.1)),
    y = 0.1,
    data = post1,
    breaks = 50,
    fill = alpha(desc_color, 0.35),
    color = desc_color_dark,
    side = "bottom",
    .width = .95,
    scale = 1,
    height = 3
  ) +
  coord_cartesian(clip = "off", ylim = c(0.7, nrow(post1_samples_rope)))

p_rope_densities + p_rope_props + plot_annotation(tag_levels = "A")
```

@@

```{r}
#| label: fig-p-neg
#| fig-width: 2.7
#| fig-height: 2.2
#| fig-cap: "Bayesian derivation of the posterior distribution of $p^- = Pr(\\textrm{Valence effect} \\le 0)$, the proportion of individuals with negative valence effects on logRT. Thin gray lines show 100 CDFs of the distribution of valence effects calculated from random draws of $\\beta_1$ and $\\tau_1$, as in @fig-2 B. Blue lines show the calculation of the posterior distribution of $p^-$, which is also depicted as a marginal histogram at the top left. The point estimate and interval represents the posterior mean and 95\\%CI. Compare to the point estimate of $p^-$ in @fig-2 B."

# Add summaries of CDF < 0 to CDF curves as inset plots
p_cdf_samples_with_p_neg <- p_cdf_samples +
  geom_segment(
    aes(x = 0, xend = 0, y = 0, yend = cdf(dist_valence, 0)),
    data = post1_samples,
    color = desc_color,
    alpha = 0.05
  ) +
  geom_segment(
    aes(x = 0, xend = I(0), y = cdf(dist_valence, 0), yend = cdf(dist_valence, 0)),
    data = post1_samples,
    color = desc_color,
    alpha = 0.1
  ) +
  stat_histinterval(
      aes(y = cdf(dist_valence, 0)),
      x = I(0),
      data = post1,
      breaks = 50,
      fill = alpha(desc_color, 0.35),
      color = desc_color_dark,
      side = "left",
      .width = .95,
      scale = 1,
      width = 0.1
  ) + 
  annotate(
    "text", 
    x = 0, y = I(0), 
    label = "0", 
    color = desc_color, 
    vjust = 1.3, 
    fontface = "bold", 
    size = 3.5
  ) +
  geom_text( 
    aes(x = I(0), y = cdf(dist_valence, 0)),
    data = post1_mean,
    label = expression(bolditalic(p ^ "−")), 
    color = desc_color, 
    hjust = 1.4, 
    fontface = "bold", 
    size = 3.5
  ) +
  scale_x_valence(breaks = c(-0.4, -0.2, 0.2)) +
  coord_cartesian(clip = "off")

p_cdf_samples_with_p_neg
```
